---
title: "STA305 Technical Report"
author: "Group 8"
date: "16/08/2021"
geometry: margin=2.5cm
output:
  pdf_document:
    latex_engine: xelatex
---
\fontfamily{Arial}
\fontsize{10}{22}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(BDEsize)
```

# Objective (Question of Interest)
* Which kernel type yields the most popcorn controlling the oil type?
* Which oil type yields the most popcorn controlling the kernel type?
* Does the yield of popcorn by kernel type vary depending on the oil used?

# Variables
Predictor variables:

* Kernel type : Baby white (A), Ladyfinger (B), Medium White (C), Midnight blue (D)

* Oil type: Canola oil (1), Sunflower (2), Safflower oil (3)

Response variable:

* Yield of popcorn kernels (measured by number of popped kernels)

Blocking variable: 

* Pot size (small, medium, large)

## Load the data
```{r}
pop_data <- read_excel("STA305_FinalProjectData.xlsx")
attach(pop_data)

table(OIL,KERNEL)
```
Every treatment combination is possible, there are 9 observations in each treatment level, so this is a completed and balanced design, with crossed factor.


# Sample Size Calculations

For the sample size calculations, we will set the probability of a type I error to be 5%, i.e., $P(\text{reject } H_0 \mid H_0 \text{ is true}) = 0.05$. As for the power, we will set the power to be the standard 80% which is to say that the probability of a type II error is set at 20%, i.e., $P(\text{fail to reject } H_0 \mid H_0 \text{ is false}) = 0.05$. Furthermore, detecting small effects requires larger sample sizes. Hence, given the time limitations of the experiment, we will consider detecting only large effects to be suffficient. Given these parameters, the sample size calculations are done in the code below. The minimum sample size required for our RCBD with an interaction term is three batches of popcorn per treatment, per block. In this case, the exact effect sizes are 0.68 for the oil type, 0.69 for the kernel type, and 0.998 for the interaction term. As per Cohen, this corresponds to a medium-large effect for the factors of interest (oil type and kernel type)and a (very) large effect for the interaction term. If interactions are removed from the model, the minimum sample size is reduced to two batches per treatment, per block. In this case, the effect sizes are 0.85 for oil type and 0.87 for the kernel type. Given these results, we will go with a sample size of three batches of popcorn per treatment, per block. Since the oil type factor has three levels and the kernel type factor has four levels, there are 12 treatments. Since each block will get all treatments (and there are three blocks), the total sample size is 108 batches of popcorn (12 treatments times three blocks times three batches per group).

```{r}
# With interaction
Size.Block(factor.lev=c(3, 4), interaction=TRUE, delta_type=1, delta=c(1, 1, 1), 
  alpha=0.05, beta=0.2)
# Without interaction
Size.Block(factor.lev=c(3, 4), interaction=FALSE, delta_type=1, delta=c(1, 1, 1), 
  alpha=0.05, beta=0.2)
```

# Model 1: Interaction and main effect model
$$
\begin{aligned}
   Y_{ijk}=\mu + \alpha_{j} + \beta_{k}+ (\alpha_{j}*\beta_{k}) + \gamma_{g} + e_{ijk}
\end{aligned}
$$

* $Y_{ijk}$ is the i-th observation from the j-th kernel type and k-th oil type;
* $\mu$ is the grand mean of the popcorn yield;
* $\alpha_{j}$ is the deviation of the mean of j-th kernel type from the grand mean;
  j=1,2,3,4;
* $\beta_{k}$ is the deviation of the mean of the k-th oil type from the grand mean; 
  k=1,2,3
* $\alpha_{j}*\beta_{k}$ is the interaction effect between kernel type and oil type;
* $\gamma_{g}$ is the deviation of the mean of the g-th block (pot size) from the grand mean;
  g=1,2,3,4;
* $e_{ijk}$ is random error. We assume $e_{ijk}$ ~ $N(0,\sigma^{2})$

## Dummy Coding Parametrization
$$
\begin{aligned}
y_{i} &= \beta_{0}+\beta_{1}*I_{B,i}+\beta_{2}*I_{C,i} + \beta_{3}*I_{D,i} + \beta_{4}*I_{2,i} + \beta_{5}*I_{3,i} \\
&+ \beta_{6}*[I_{B,i}*I_{2,i}] + \beta_{7}*[I_{C,i}*I_{2,i}] + \beta_{8}*[I_{D,i}*I_{2,i}] + \beta_{9}*[I_{B,i}*I_{3,i}] \\
&+ \beta_{10}*[I_{C,i}*I_{3,i}] + \beta_{11}*[I_{D,i}*I_{3,i}] + e_{i}
\end{aligned}
$$

## Hypothesis:
$$
\begin{aligned}
H_{0} &:\beta_{5}=\beta_{6}=\beta_{7}=\beta_{8}=\beta_{9}=\beta_{10}=\beta_{11}=0 \\
H_{a} &:\ at\ least\ one\ \beta_{k} \ne 0,\ for\ k=5,6,...,11
\end{aligned}
$$

```{r}
model_1 <- lm(NUM_POPPED~POT+KERNEL*OIL)
anova(model_1)
```
## Test Statistic
$$
\begin{aligned}
F &= \frac{MSab}{MSE} = 0.4677 \sim F_{6,94}\\
p &= P (F_{6,94}>F)=0.8307 > 0.05
\end{aligned}
$$

## Conclusion
Since p-value is larger than our $\alpha$, significance level, we fail to reject $H_0$. Therefore, there is insufficient evidence that popcorn yield by kernel types vary depending on the type of oil used when making popcorn. Hence, interaction was not statistically significant and we will thus need to fit an additive model.

# Model 2: Additive model
$$
\begin{aligned}
   Y_{ijk}=\mu + \alpha_{j} + \beta_{k} + \gamma_{g}+e_{ijk}
\end{aligned}
$$

* $Y_{ijk}$ is the i-th observation from the j-th kernel type and k-th oil type;
* $\mu$ is the grand mean of the popcorn yield;
* $\alpha_{j}$ is the deviation of the mean of j-th kernel type from the grand mean;
  j=1,2,3,4;
* $\beta_{k}$ is the deviation of the mean of the k-th oil type from the grand mean; 
  k=1,2,3;
* $\gamma_{g}$ is the deviation of the mean of the g-th block (pot size) from the grand mean;
  g=1,2,3,4;
* $e_{ijk}$ is random error. We assume $e_{ijk}$ ~ $N(0,\sigma^{2})$


## Dummy Coding Parametrization
$$
\begin{aligned}
y_{i} = \beta_{0}+\beta_{1}*I_{B,i}+\beta_{2}*I_{C,i} + \beta_{3}*I_{D,i} + \beta_{4}*I_{2,i} + \beta_{5}*I_{3,i} + e_{i}
\end{aligned}
$$

## Hypothesis for Kernel Type
$$
\begin{aligned}
H_{0} &:\beta_{1}=\beta_{2}=\beta_{3}=0 \\
H_{a} &:\ at\ least\ one\ \beta_{k} \ne 0,\ for\ k=1,2,3
\end{aligned}
$$

## Hypothesis for Oil Type
$$
\begin{aligned}
H_{0} &:\beta_{4}=\beta_{5}=0 \\ 
H_{a} &:\ at\ least\ one\ \beta_{k} \ne 0,\ for\ k=4,5
\end{aligned}
$$

```{r}
model_2 <- lm(NUM_POPPED~POT + KERNEL+OIL)
anova(model_2)
```
## Test Statistic

### Kernel Effect
$$
\begin{aligned}
F &= \frac{MSkernel}{MSE} = 18.9785 \sim F_{3,100}\\
p &= P (F_{3,100}>F)= 8.058\times10^{-10} < 0.05
\end{aligned}
$$
  
### Oil effect
$$
\begin{aligned}
F &= \frac{MSoil}{MSE} = 1.1098 \sim F_{2,100}\\
p &= P (F_{2,100}>F)= 0.3336 > 0.05
\end{aligned}
$$

## Conclusion

### Kernel Effect
Since p-value is smaller than our $\alpha$, significance level, we reject $H_0$. Therefore, we have strong evidence to conclude that kernel type affects popcorn yield when controlling for oil.

### Oil effect
Since p-value is larger than our $\alpha$, significance level, we fail to reject $H_0$. Hence, we have inconclusive evidence to conclude that popcorn yield varies by oil type.


# Checking necessary assumptions
```{r, out.width="60%", fig.align='center'}
par(mfrow=c(2,2))

plot(model_1, 2) #normality of model 1
plot(model_2, 2) #normality of model 2

plot(model_1,1) #homogeneity of variance model 1
plot(model_2,1) #homogeneity of variance model 2
```
* Normality of the observations: according to both the QQ plot and the Shapiro-Wilk normality test, p= 0.00037. There is sifficient evidence that the observations are not normally distributed. Therefore, the assumption of normality is violated. This is a limitation of our analysis that we cannot make inference to the population.

* Independence of observation: this assumption is checked by the nature of our experimental design. Each batch of kernel is only given one treatment, and the pots containing the kernels and oil were washed after every single trial, so the result of one batch does not depend on the result of the other batch.

* Homogeneity within blocks: this assumption is checked by the nature of our experimental design. Our blocking variable is the pot size in small, medium and large. The one pot is used to match with one size.

* Heterogeneity between blocks: this assumption is checked by the nature of our experimental design. Our blocking variable is the pot size in small, medium and large. 3 different pots in 3 distinct sizes were used.

* Homogeneity of variance between treatment: according to the residual versus fit graph above, the dots are scattered fairly evenly on the upper- and lower-side of the red line. Therefore, homogeneity of variance is checked.



# Post Hoc Analysis

## Tukey's HSD Procedure
```{r}
tukeyCIs <- TukeyHSD(aov(model_2),factor=KERNEL,
conf.level = 0.95)
tukeyCIs
```

```{r, fig.dim = c(8, 2), fig.align='center'}
par(mfrow=c(1, 3))
plot(tukeyCIs)
```
# Conclusion
