---
title: "STA305 Technical Report"
author: "Group 8"
date: "16/08/2021"
output:
  pdf_document:
    latex_engine: xelatex
---
\fontfamily{Arial}
\fontsize{12}{22}
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

### Objective (Question of Interest)
* Which kernel type yields the most popcorn controlling the oil type?
* Which oil type yields the most popcorn controlling the kernel type?
* Does the yield of popcorn by kernel type vary depending on the oil used?

Variables
Predictor variables:
  Kernel type : Baby white (A), Ladyfinger (B), Medium White (C), Midnight blue (D)
  Oil type: Canola oil (1), Sunflower (2), Safflower oil (3)
Response variable:
  Yield of popcorn kernels (measured by number of popped kernels)
Blocking variable: Pot size (small, medium, large)

#### Load the data
```{r}
pop_data <- read_excel("STA305_FinalProjectData.xlsx")
attach(pop_data)

table(OIL,KERNEL)
```

Every treatment combination is possible, there are 9 observations in each treatment level, so this is a completed and balanced design, with crossed factor.

## Model 1: Interaction and main effect model
  $y_{i} = \beta_{0}+\beta_{1}*I_{B,i}+\beta_{2}*I_{C,i} + \beta_{3}*I_{D,i}$
  $+ \beta_{4}*I_{2,i} + \beta_{5}*I_{3,i} + \beta_{6}*[I_{B,i}*I_{2,i}]$
  $+ \beta_{7}*[I_{C,i}*I_{2,i}] + \beta_{8}*[I_{D,i}*I_{2,i}] +$
  $\beta_{9}*[I_{B,i}*I_{3,i}] + \beta_{10}*[I_{C,i}*I_{3,i}] +$
  $\beta_{11}*[I_{D,i}*I_{3,i}] + e_{i}$

### Hypothesis
  $H_{0}:\beta_{5}=\beta_{6}=\beta_{7}=\beta_{8}=\beta_{9}=\beta_{10}=\beta_{11}=0$
  
  $H_{a}:\ at\ least\ one\ \beta_{k} \ne 0,\ for\ k=5,6,...,11$

### ANOVA
```{r}
model_1 <- lm(NUM_POPPED~KERNEL*OIL)
anova(model_1)

model_2 <- lm(NUM_POPPED~POT+KERNEL*OIL)
anova(model_2)

```

  $\gamma_{g}$
###where:
   $y_{i}\ is\ $ 